### 2016-07-03 ansi-isolation.tex
Typos fixed. ANSI ISO figure split.

\begin{figure*}[t!]
\begin{smathpar}
\begin{array}{lcl}
(\A,\visZ) \Vdash \eta \in T_i & \defeq & \eta \in \A \conj \txn(\eta) = T_i\\
(\A,\visZ) \Vdash \eta_1 \visar \eta_2 & \defeq & \{\eta_1,\eta_2\}
        \subseteq \A \conj (\eta_1,\eta_2) \in \visZ\\
(\A,\visZ) \Vdash \eta_1 \soar \eta_2 & \defeq & \{\eta_1,\eta_2\}
        \subseteq \A \conj \txn(\eta_1)=\txn(\eta_2) \conj \id(\eta_1)
        < \id(\eta_2)\\
\E \Vdash S \subseteq T_i & \defeq & \forall \eta.~ \eta
        \in S \Rightarrow \underE{\eta \in T_i} \\
\E \Vdash T_i \visar \eta & \defeq &\forall\eta_1
        .\,(\E \Vdash \eta_1 \in T_i) \Rightarrow \E \Vdash \eta_1 \visar \eta \\
\underE{T_i \visar T_j} & \defeq &  \forall\eta_1,\eta_2.\,
    %\sameobj{\eta_1}{\eta_2}  \Rightarrow 
    \underE{\eta_1\in T_i} \conj \underE{\eta_2 \in T_j} \Rightarrow 
    \underE{\eta_1 \visar \eta_2} \\
\underE{T_i \invisar T_j} & \defeq &  \forall\eta_1,\eta_2.\,
        %\sameobj{\eta_1}{\eta_2}\Rightarrow 
        \underE{\eta_1\in T_i} \conj \underE{\eta_2 \in T_j} \Rightarrow 
        \neg (\underE{\eta_1 \visar \eta_2})\\
\underE{T_i \wrstoar X} & \defeq & \exists\eta.~
        \underE{\eta \in T_i} \conj \kind(\eta) = \C{WR}(X)\\
\underE{T_i \rdsfmar X} & \defeq & \exists\eta.~
        \underE{\eta \in T_i} \conj \kind(\eta) = \C{RD}(X)\\
\underE{T_i \usesar X} & \defeq & \underE{T_i \wrstoar X} \disj
      \underE{T_i \rdsfmar X}\\
\underE{\C{RMWVis}(T_j)} & \defeq & \forall\eta_1,\eta_2.\,
       \underE{\{\eta_1,\eta_2\} \subseteq T_j} \conj
       \underE{\eta_1 \soar \eta_2} \Rightarrow \underE{\eta_1 \visar
       \eta_2}\\
\underE{\C{MonotonicVis}(T_j)} & \defeq & 
       \underE{\C{RMWVis}(T_j)} \conj 
       \forall\eta,\eta_1,\eta_2.\, \underE{\{\eta_1,\eta_2\} \in T_j} 
          \conj \\
  &   & \hspace*{1.2in}\underE{\eta \visar \eta_1} \conj
        \underE{\eta_1 \soar \eta_2} \Rightarrow \underE{\eta \visar
        \eta_2} \\
%  \C{CausalVis}(T_i) & \defeq & 
%         \C{MonotonicVis}(T_i) \conj \C{RMWVis}(T_i)\\
\underE{\C{AtomicVis}(T_j)} & \defeq & 
       \forall\eta_1,\eta_2.\, \neg(\underE{\eta_1 \in T_j}) \conj
       \underE{\eta_2 \in T_j} \conj
       \underE{\eta_1 \visar \eta_2} \Rightarrow \underE{\txn(\eta_1)
       \visar \eta_2}\\
\underE{\C{CommitVis}(T_j)} & \defeq & 
       \forall\eta_1,\eta_2.~ \neg(\underE{\eta_1 \in T_j}) \conj 
          \underE{\eta_2 \in T_j} \conj
       \underE{\eta_1 \visar \eta_2} \Rightarrow\\
  &   & \hspace*{1.2in}\exists\eta.\, \underE{\eta \in \txn(\eta_1)} 
        \conj \kind(\eta) = \C{COMMIT} 
        \conj \underE{\eta \visar \eta_2}\\
% \underE{\C{TransVis}(T_j)} & \defeq &  \forall
%        \eta_1,\eta_2,\eta_3.\, \underE{\eta_3 \in T_j} \conj
%        \underE{\eta_1 \visar \eta_2} \conj
%        \underE{\eta_2 \visar \eta_3} \Rightarrow \underE{\eta_1 \visar
%        \eta_3} \\
\underE{\C{RC}(T_j)} & \defeq & \underE{\C{AtomicVis}(T_j)} 
        \conj \underE{\C{CommitVis}(T_j)}\\
\underE{\C{MAV}(T_j)} & \defeq & \underE{\C{RC}(T_j)} \conj
      \underE{\C{MonotonicVis}(T_j)} \\
\underE{\C{SnapshotVis}(T_i,T_j)} & \defeq &  \underE{T_i
       \visar T_j} \disj \underE{T_i \invisar T_j}\\
\underE{\C{RR}(T_j)} & \defeq & \underE{\C{MAV}(T_j)}
       \conj \forall T_i.\,T_i \neq T_j \Rightarrow 
        \underE{\C{SnapshotVis}(T_i,T_j)} \\
% &   & \hspace*{2in}\conj  \C{SnapshotVis}(T_i,T_j)\\
\underE{\C{SnapshotSER}(T_i,T_j,X)} & \defeq &  \underE{T_i
       \visar T_j} \disj (\underE{T_i \invisar T_j} \conj \\
  &   &\hspace*{1.2in}\exists \eta.~\underE{\eta\in T_i} \conj \kind(\eta) =
       \C{WR(X)} \Rightarrow \underE{T_j \visar \eta})\\
\underE{\C{SI}(T_j)} & \defeq &  \underE{\C{RR}(T_j)}
       \conj \forall T_i.\,(T_i \neq T_j \conj \exists X.\, \underE{T_i \wrstoar X} \conj 
        \underE{T_j \wrstoar X})\\
%      (\exists X.\, T_i \wrstoar X \conj T_j \wrstoar X)
%      \Rightarrow  \C{TotalVis}(T_i,T_j)\\
  &   & \hspace*{2in} \Rightarrow \underE{\C{SnapshotSER}(T_i,T_j,X)}\\
\underE{\C{SER}(T_j)} & \defeq & \underE{\C{RR}(T_j)}
        \conj \forall T_i.\,(T_i \neq T_j \conj \exists X.\, 
        \underE{T_i \wrstoar X} \conj \underE{T_j \usesar X})\\
  &   & \hspace*{2in} \Rightarrow \underE{\C{SnapshotSER}(T_i,T_j,X)}\\
\end{array}
\end{smathpar}

\caption{Standard isolation guarantees expressed as trace
well-formedness constraints}
\label{fig:ansi-isolation}
\end{figure*}

Fig.~\ref{fig:ansi-isolation} shows the specification of standard
isolation guarantees expressed as constraints over trace
well-formedness. For brevity and convenience, we adopt few notations
in this and subsequent sections.  An execution trace is destructed
into $\A$ and $\visZ$ whenever individual components of the pair are
needed. Otherwise, it is written as $\E$. Sometimes, the dot notation
(\eg~$\E.\A$) is also used. Since $\A$ and $\visZ$ are both sets, we
lift the operations on sets to pairs of sets when updating $\E$. For
example, $\E' = \E \cup (\{\eta_2\},\{(\eta_1,\eta_2)\})$ expands to
$\E' = (\E.\A \cup \{\eta_2\},\,\E.\visZ \cup \{(\eta_1,\eta_2)\})$.
When $\psi$ is a formula, $\underE{\psi}$ denotes the
interpretation of $\psi$ in the context of the trace $\E$. Such
interpretations are defined on a case-by-case basis in
Fig.~\ref{fig:ansi-isolation}. In the following, we give informal
explanations for each definition.

In the context of a trace $(\A,\visZ)$, an effect $\eta$ is said to
belong to a transaction $T_i$ if $\eta$ belongs to the effect set $A$
and its transaction identifier is $T_i$. The containment relation is
trivially lifted to the set of effects to define $\underE{S \subseteq
  T_i}$.  The interpretation of $\eta_1 \visar \eta_2$ and $\eta_1
\soar \eta_2$ are as explained earlier.  A transaction $T_i$ is said
to be visible to an effect $\eta$ if every effect $\eta_1$ of $T_i$
recorded by the trace is visible to $\eta$.  $T_i$ may be visible to
$\eta$ but may not be visible to every other effect in the
transaction. For a transaction $T_i$ to be considered to be visible to
a transaction $T_j$ in the context of a trace $\E$ (written
$\underE{T_i \visar T_j}$), every effect ($\eta_1$) of $T_i$ present
in $\E$ must be visible to every effect ($\eta_2$) of $T_j$ in
$\E$. Conversely, if none of the effects of $T_i$ present in $\E$ are
visible to any effect of $T_j$, then $T_i$ is considered invisible to
$T_j$ under $\E$ (written $\underE{T_i \invisar T_j}$). Transaction
$T_i$ is said to have written to a variable $X$ under $\E$ (i.e.,
$\underE{T_i \wrstoar X}$) if there exists a write-to-$X$ effect from
$T_i$ in $\E$. \emph{Reads-from} is similar. $T_i$ \emph{uses} $X$ if
it reads or writes $X$.

Various isolation guarantees are defined as propositions indexed by
a transaction identifier $T_j$. Transaction $T_j$ is said to have
experienced \emph{read-my-writes} visibility under $\E$ if every
effect ($\eta_1$) of $T_j$ is visible to every subsequent effect
($\eta_2$) of the same transaction in $\E$. This lets $T_j$ to never
lose its own updates. Monotonic visibility adds one more constraint to
read-my-writes; besides requiring $\eta_1$ to be visible to $\eta_2$,
it also requires every effect ($\eta$) visible to $\eta_1$ in $\E$ to
be visible to $\eta_2$ as well. Thus $T_j$ witnesses monotonically
increasing state as time progresses. Atomic visibility allows an
effect $\eta_2$ of $T_j$ to witness an effect $\eta_1$ of $T_i$ only
if all effects of $T_i$ in $\E$ are also visible to $\eta_2$. Atomic
visibility thus prevents a transaction from being partially visible.
However, atomic visibility does not prevent an uncommitted
transaction from being visible. This is addressed by \C{CommitVis},
which requires the commit effect of $T_i$ to be visible whenever any
effect of $T_i$ is visible.

The ANSI SQL 92 standard requires \iso{Read Committed} isolation to avoid
the dirty reads phenomenon, which is achieved by enforcing
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
specification is therefore a combination of these two guarantees. The
specification also agrees with the description and
implementation~\cite{bailishat,pldi15} of {\sc rc} for highly
available replicated stores. On relational databases, however, {\sc
rc} has also come to be associated with the $\mathtt{MonotonicVis}$
guarantee.  Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$
are sufficient to reason about {\sc rc} isolation on relational stores
too. The combination of these guarantees with the {\sc sc} property of
relational stores (formalized in \S\ref{sec:store-consistency})
automatically leads to the monotonicity guarantee, which explains why
{\sc rc} comes with $\mathtt{MonotonicVis}$ on such stores regardless
of the implementation. On weakly consistent stores however,
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ do not imply
$\mathtt{MonotonicVis}$. A stronger isolation level called
\iso{Monotonic Atomic View} ({\sc mav} of
Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
explicitly extend {\sc rc} with monotonicity on such stores. The sample
{\sc rc} executions shown in Fig.~\ref{fig:rc-executions} are also
{\sc mav}.

\begin{figure}
\centering
\subcaptionbox {
  {\sc rr}($T_1$), {\sc mav}($T_2$).
  \label{fig:ansi-iso-eg-rr}
} [
  0.33\columnwidth
] {
  \includegraphics[scale=0.36]{Figures/rr-eg}
}
\subcaptionbox {
  {\sc si}($T_1$), {\sc mav}($T_2$)
  \label{fig:ansi-iso-eg-si}
} [
  0.33\columnwidth
] {
  \includegraphics[scale=0.36]{Figures/si-eg-2}
}
\subcaptionbox {
  {\sc ser}($T_3$), {\sc mav}($T_4$)
  \label{fig:ansi-iso-eg-ser}
}{
  \includegraphics[scale=0.32]{Figures/ser-eg}
}

@@

A transaction $T_i$ is said to be snapshot-visible to a transaction
$T_j$ if either it is visible to $T_j$, or it is invisible; it is
forbidden for only a suffix (more generally, a subset) of $T_j$ to
witness $T_i$.  \iso{Repeatable Read} ({\sc rr})
isolation extends {\sc mav} with the snapshot visibility guarantee.
Observe that snapshot visibility permits $T_i$ and $T_j$ to execute
and commit while being oblivious of each other. This scenario is
captured in Fig.~\ref{fig:ansi-iso-eg-rr}, where transactions $T_1$
and $T_2$, which perform conflicting writes, execute against a
snapshot of the database and commit concurrently.

\iso{Snapshot Isolation} ({\sc SI}) proscribes this possibility. If
$T_i$ and $T_j$ both write to the same shared variable, then {\sc si}
insists that either $T_i$ be visible to $T_j$, or $T_j$ be visible to
the conflicting write of $T_i$ (this is captured by the auxiliary
definition \C{SnapshotSER}). Fig.~\ref{fig:ansi-iso-eg-si} shows an
execution where {\sc si} transaction $T_1$ witnesses a snapshot of the
database that doesn't include $T_2$, but the conflicting write to $X$
in $T_2$, and the subsequent operations (due to the monotonicity
requirement of {\sc mav}) witness $T_1$. Thus, $T_2$ is aware of $T_1$
when it commits. Lastly, \iso{Serializable} isolation extends
$\mathtt{SnapshotSER}$ to also cover non-conflicting transactions that
write to variables read in the current transaction.
Fig.~\ref{fig:ansi-iso-eg-ser} shows an execution of an {\sc ser}
transaction $T_3$ and a {\sc mav transaction} $T_4$, which write to $X$
and $Y$, respectively. The snapshot witnessed by the {\sc ser}
transaction $T_3$ does not include $T_4$, but the write to $Y$ in
$T_4$, although non-conflicting, witnesses $T_1$ because $Y$ is read
by $T_1$. Visibility includes other operations of $T_4$ because of
{\sc mav}.

@@

% \iso{Snapshot Isolation} spec ({\sc si}) extends {\sc rr} with a
% one-way serializability guarantee w.r.t the transactions that perform
% conflicting writes (i.e., writes to the same shared variable).
% \footnote{Fig.~\ref{fig:ansi-isolation} presents slightly
% weaker versions of {\sc si} and {\sc ser} specs in the interest of
% clarity.}. Fig.~\ref{fig:ansi-iso-eg} shows sample executions of
% transactions $T_1$ and $T_2$. Both transactions read and write to
% shared variables $X$ and $Y$. In the first execution, $T_2$ commits
% while $T_1$ is still in progress, but {\sc rr} isolation prevents
% $T_1$ from witnessing the writes of $T_2$. In the second execution,

>>>>

% ISOLATION SPECS
% ---------------
\input{iso-specs}

We now describe specifications of standard isolation guarantees
expressed as constraints over trace well-formedness. For brevity and
convenience, we adopt few notations in this and subsequent sections.
An execution trace is destructed as ($\A$,$\visZ$) whenever individual
components of the pair are needed. Otherwise, it is written as $\E$.
Sometimes, the dot notation (\eg~$\E.\A$) is also used. Since $\A$ and
$\visZ$ are both sets, we lift the operations on sets to pairs of sets
when updating $\E$. For example, $\E' = \E \cup
(\{\eta_2\},\{(\eta_1,\eta_2)\})$ expands to $\E' = (\E.\A \cup
\{\eta_2\},\,\E.\visZ \cup \{(\eta_1,\eta_2)\})$.  When $\psi$ is a
formula, $\underE{\psi}$ denotes the interpretation of $\psi$ in the
context of the trace $\E$. Such interpretations are defined on a
case-by-case basis in Figs.~\ref{fig:rel-defs}
and~\ref{fig:ansi-isolation}.

Fig.~\ref{fig:rel-defs} shows various relations defined over elements
in a trace. In the context of a trace $(\A,\visZ)$, an effect $\eta$
is said to belong to a transaction $T_i$ if $\eta$ belongs to the
effect set $A$ and its transaction identifier is $T_i$. The
containment relation is trivially lifted to the set of effects to
define $\underE{S \subseteq T_i}$.  Visibility and session order
relations are denoted by $\eta_1 \visar \eta_2$ and $\eta_1 \soar
\eta_2$, respectively. A transaction $T_i$ is said to be visible to an
effect $\eta$ if every effect $\eta_1$ of $T_i$ recorded by the trace
is visible to $\eta$.  $T_i$ may be visible to $\eta$ but may not be
visible to every other effect in the $\txn(\eta)$. For a transaction
$T_i$ to be considered to be visible to a transaction $T_j$ in the
context of a trace $\E$ (written $\underE{T_i \visar T_j}$), every
effect ($\eta_1$) of $T_i$ present in $\E$ must be visible to every
effect ($\eta_2$) of $T_j$ in $\E$.  Conversely, if none of the
effects of $T_i$ present in $\E$ are visible to any effect of $T_j$,
then $T_i$ is considered invisible to $T_j$ under $\E$ (written
$\underE{T_i \invisar T_j}$). Transaction $T_i$ is said to have
written to a variable $X$ under $\E$ (i.e., $\underE{T_i \wrstoar X}$)
if there exists a write-to-$X$ effect from $T_i$ in $\E$.
\emph{Reads-from} ($\underE{T_i \rdsfmar X}$) is defined similarly.
$T_i \usesar X$ if it reads or writes $X$.

Fig.~\ref{fig:ansi-isolation} shows various isolation guarantees
defined as propositions indexed by transaction identifiers.
Transaction $T_j$ is said to have experienced \emph{read-my-writes}
visibility under $\E$ if every effect ($\eta_1$) of $T_j$ is visible
to every subsequent effect ($\eta_2$) of the same transaction in $\E$.
This lets $T_j$ to never lose its own updates. Monotonic visibility
adds one more constraint to read-my-writes; besides requiring $\eta_1$
to be visible to $\eta_2$, it also requires every effect ($\eta$)
visible to $\eta_1$ in $\E$ to be visible to $\eta_2$ as well. Thus,
later operations in a transaction witness at least the same set of
effects witnessed by the earlier operations, if not more (hence,
``monotonic'' visibility). Atomic visibility allows an effect
$\eta_2$ of $T_j$ to witness an effect $\eta_1$ of $T_i$ only if all
effects of $T_i$ in $\E$ are also visible to $\eta_2$. Atomic
visibility thus prevents a transaction from being partially visible.
However, atomic visibility does not prevent an uncommitted transaction
from being visible. This is addressed by \C{CommitVis}, which requires
the commit effect of $T_i$ to be visible whenever any effect of $T_i$
is visible.

The ANSI SQL 92 standard requires \iso{Read Committed} isolation to
avoid the dirty reads phenomenon, which is achieved by enforcing
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
specification\footnote{Note the distinction in fonts: {\sc rc} is an
abbreviation of \iso{Read Committted}, whereas \C{RC} is its
specification.} (\C{RC}) is therefore a combination of these two
guarantees. The specification also agrees with the description and
implementation~\cite{bailishat,pldi15} of {\sc rc} for highly
available replicated stores. On relational databases, however, {\sc
rc} has also come to be associated with the $\mathtt{MonotonicVis}$
guarantee.  Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$
are sufficient to reason about {\sc rc} isolation on relational stores
too. The combination of these guarantees with the {\sc sc} property of
relational stores (formalized in \S\ref{sec:store-consistency})
automatically leads to the monotonicity guarantee, which explains why
{\sc rc} comes with $\mathtt{MonotonicVis}$ on such stores regardless
of the implementation. On weakly consistent stores however,
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ do not imply
$\mathtt{MonotonicVis}$. A stronger isolation level called
\iso{Monotonic Atomic View} ({\sc mav} of
Fig.~\ref{fig:ansi-isolation})~\cite{bailishat} was proposed to
explicitly extend {\sc rc} with monotonicity on such stores.

\begin{figure}
\centering
\subcaptionbox {
  {\sc rr}($T_1$), {\sc rr}($T_2$).
  \label{fig:ansi-iso-eg-rr}
} [
  0.33\columnwidth
] {
  \includegraphics[scale=0.36]{Figures/rr-eg}
}
\subcaptionbox {
  {\sc si}($T_1$), {\sc mav}($T_2$)
  \label{fig:ansi-iso-eg-si}
} [
  0.33\columnwidth
] {
  \includegraphics[scale=0.36]{Figures/si-eg-2}
}
\subcaptionbox {
  {\sc ser}($T_3$), {\sc mav}($T_4$)
  \label{fig:ansi-iso-eg-ser}
}{
  \includegraphics[scale=0.32]{Figures/ser-eg}
}

@@

Snapshot visibility (\C{SnapshotVis}) captures the scenario where a
transaction executes against a static snapshot of the database.  A
transaction $T_i$ is said to be snapshot-visible to a transaction
$T_j$ if either it is visible to $T_j$ (i.e., $T_i$ is included in the
snapshot), or it is invisible (i.e., it is not included); it is
forbidden for only a suffix (more generally, a subset) of $T_j$ to
witness $T_i$. \iso{Repeatable Read} ({\sc rr}) isolation extends {\sc
mav} with the snapshot visibility guarantee.  Observe that snapshot
visibility permits $T_i$ and $T_j$ to execute and commit while being
oblivious of each other. This scenario is captured in
Fig.~\ref{fig:ansi-iso-eg-rr}, where transactions $T_1$ and $T_2$,
which perform conflicting writes, execute against a snapshot of the
database and commit concurrently. While the actual values read and
written by $T_1$ and $T_2$ are unimportant (hence, elided), it is
important to note the \emph{absence} of visiblity arrows from $T_2$ to
$T_1$, although $T_2$ commits before $T_1$'s read-from-$Y$.

\iso{Snapshot Isolation} ({\sc si}) proscribes this possibility. If
$T_i$ and $T_j$ both write to the same shared variable, then {\sc si}
insists that either $T_i$ be visible to $T_j$, or $T_j$ be visible to
the conflicting write of $T_i$ (this is captured by the auxiliary
definition \C{SnapshotSER}). Fig.~\ref{fig:ansi-iso-eg-si} shows an
execution where {\sc si} transaction $T_1$ witnesses a snapshot of the
database that doesn't include $T_2$, but the conflicting write to $X$
in $T_2$, and the subsequent operations (due to the monotonicity
requirement of {\sc mav}) witness $T_1$. Thus, $T_2$ is aware of $T_1$
when it commits. Lastly, \iso{Serializable} isolation extends
$\mathtt{SnapshotSER}$ to also cover non-conflicting transactions that
write to variables read in the current transaction.
Fig.~\ref{fig:ansi-iso-eg-ser} shows an execution of an {\sc ser}
transaction $T_3$ and a {\sc mav transaction} $T_4$, which write to $X$
and $Y$, respectively. The snapshot witnessed by the {\sc ser}
transaction $T_3$ does not include $T_4$, but the write to $Y$ in
$T_4$, although non-conflicting, witnesses $T_1$ because $Y$ is read
by $T_1$. Visibility includes other operations of $T_4$ because of
{\sc mav}.

@@

% \iso{Snapshot Isolation} spec ({\sc si}) extends {\sc rr} with a
% one-way serializability guarantee w.r.t. the transactions that perform
% conflicting writes (i.e., writes to the same shared variable).
% \footnote{Fig.~\ref{fig:ansi-isolation} presents slightly
% weaker versions of {\sc si} and {\sc ser} specs in the interest of
% clarity.}. Fig.~\ref{fig:ansi-iso-eg} shows sample executions of
% transactions $T_1$ and $T_2$. Both transactions read and write to
% shared variables $X$ and $Y$. In the first execution, $T_2$ commits
% while $T_1$ is still in progress, but {\sc rr} isolation prevents
% $T_1$ from witnessing the writes of $T_2$. In the second execution,
